---
title: "TF Array Design"
author: "Max Trauernicht"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    theme: journal #cerulean
    highlight: monochrome
    toc: true
    toc_float: true
    code_folding: show
  editor_options:
    chunk_output_type: console
---

# knitr document van Steensel lab

# TF activity reporter array

# Introduction
The goal of this script is to select TF reporters from the oligo design script which can be assembled for integration in mESCs. Assembly will be performed based on bridging ligation. For this optimal bridging oligos have to be designed in this script.

# Data processing
## Path, Libraries, Parameters and Useful Functions
```{r setup, echo=FALSE, message=FALSE, warnings=FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(ggplot2)
library(seqinr)
library(seqLogo)
library(universalmotif)
library(Biostrings)
#library(SimRAD)
library(gtools)
library(DNABarcodes)
library(phylotools)
library(ape)
library(magrittr)
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(heatmaply)
library(pheatmap)
library(tibble)
library(ggseqlogo)
library(RColorBrewer)
```

### Custom functions
Functions used thoughout this script.
```{r}
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}

# Function to substring the right part of the motif
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

# Function to load PWM matrix
get_pwm_feature_matrix <- function(motif_meta_fn, fimo_fn, db = 2) {

  # validate args
  valid_dbs <- 1:2
  if(!db %in% valid_dbs)
    stop('Invalid db (database version). Please use db=1 (maintained for backward compatibility only) or db=2')

  # db=1 is maintained for backward compatibility only
  if(db == 1) {

    # read in motif metadata
    motif_meta    <- read.csv(motif_meta_fn)

    # check whether motif metadata contain essential annotations
    if(!all(c('PWM.ID', 'Cognate.TF') %in% colnames(motif_meta))) {
      message('The motif metadata file does not contain the essential columns PWM.ID and Cognate.TF')
    }

    motif_minimal <- motif_meta[, c('PWM.ID', 'Cognate.TF')]

    # load fimo output --> extract motif id, sequence id and p-value
    df <- read.table(fimo_fn)
    df <- df[, c(1, 2, 7)]

    colnames(df) <- c('PWM.ID', 'seqid', 'pval')

    # add TF id
    df <- merge(df, motif_minimal, by = 'PWM.ID')

    # group motif hits by sequence id
    l <- split(df, df[['seqid']])

    # multiple PWM and multiple hits possible. Reduce hits to one per TF, keeping best p-val only
    l <- lapply(l, function(x) {
      x_by_tf <- split(x, x[['Cognate.TF']], drop = TRUE)
      x_by_tf <- lapply(x_by_tf, function(y) y[which.min(y$pval), ])
      do.call('rbind', x_by_tf)
    })

    # initialize feature matrix
    n_tf          <- motif_minimal[['Cognate.TF']] %>%
      unique %>%
      length
    n_seq         <- length(l)
    pwm           <- matrix(1, nrow = n_seq, ncol = n_tf)
    colnames(pwm) <- (motif_minimal[['Cognate.TF']] %>% unique)

    # replace :: from names of composite motifs
    colnames(pwm) <- str_replace_all(colnames(pwm), '::', '_')

    # fill in feature matrix
    for(i in 1 : n_seq) {
      pwm[i, l[[i]][['Cognate.TF']]] <- l[[i]]$pval
    }

    # -log10 transform
    pwm           <- -1 * log10(pwm)

    # coerce to tib and return
    tib_fimo <- as_data_frame(pwm) %>%
      mutate(id = names(l))
      dplyr::select(id, everything())

  }

  # db = 2 (default)
  else {

    # load metadata
    tib_meta    <- read_csv(motif_meta_fn) %>%
      # extract tf symbol from motif id (Cognate_TF unsafe, it can be empty) and replace :: occurrences
      mutate(tf_symbol = str_remove(ID, '_[0-9]*'),
             tf_symbol = str_replace(tf_symbol, '::', '_')) %>%
      select(motif_id = `PWM ID`, tf_symbol)

    # load fimo results
    tib_fimo <- read_tsv(fimo_fn) %>%
      # extract motif id, sequence id and p-value
      select(motif_id, sequence_name, pval = `p-value`)

    # add tf symbol to fimo results
    tib_fimo <- tib_fimo %>%
      left_join(tib_meta, by = 'motif_id') %>%
      # remove hits with missing motif id (composite pwms)
      filter(!is.na(tf_symbol))

    # select best hit for each motif and sequence
    tib_fimo <- tib_fimo %>%
      group_by(sequence_name, tf_symbol) %>%
      dplyr::slice(which.min(pval)) %>%
      ungroup()

    # spread into feature matrix
    tib_fimo <- tib_fimo %>%
      mutate(pval = -1 * log10(pval)) %>%
      select(-motif_id) %>%
      spread(key = tf_symbol, value = pval, fill = 0, drop = TRUE) %>%
      # perform cosmetics on the id
      mutate(id = sequence_name) %>%
      select(-c(sequence_name)) %>%
      select(id, everything())

  }

  return(tib_fimo)

}
```


## Data import
```{r}
tf.array <- read.csv("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/oligo_design/output/mt20191205_tf-array.csv")
```

# Analysis


## Selecting TF reporters
```{r}
# Select relevant TF reporters
tf.array$TF[tf.array$TF == "Nr3c1(GR)"]  <- "Gr"
tfs.selected <- c("^Stat3$", "^Trp53$", "^Random1$", "^Rara$", "^Gr$", "^Nfya$")
tf.array <- tf.array[grep(paste(tfs.selected, collapse = "|"),tf.array$TF),]
tf.array <- tf.array[tf.array$Promoter == "minP" & tf.array$Distance == "21bp" &
                       tf.array$Spacing == "10bp" & tf.array$Background == 1 &
                       tf.array$Barcode == 4,]

# Attach pA site (sNRP-1)
tf.array$pA <- "cttgtgactgggaaaaccctggcgtaaataaaatacgaaatg"

# Remove 5' and 3' primer adapter
tf.array <- tf.array %>% select(-Primer1_seq, -Primer2_seq)

# Concatenate intermediate sequence
tf.array$reporter <- paste(tf.array$motif1, tf.array$Space1, tf.array$motif2,
                           tf.array$Space2, tf.array$motif3, tf.array$Space3,
                           tf.array$motif4, tf.array$Distance_seq, tf.array$Promoter_sequence,
                           tf.array$S1_primer, tf.array$barcode, tf.array$pA, sep = "")
tf.array <- tf.array[,c(-2:-19)]
```


## Add random oligos as adapters for the bridging oligos
```{r}
## Generate 18bp oligos - select top 12 oligos
# moderate GC content
# no TF binding in front of/after
# EcoRI sites in exactly 2
## Adding 3x random 11bp 'motifs' that serve as negative control
x <- cbind.data.frame(c(1), c("GGCGTTTACTATGGGAGGTCTATATAAGCAGAGCTCGTTTAGTGAACCGTCAGATC"))
names(x) <- c("seq.name", "seq.text")


# Generate 2000 random 10bp oligos
# GC content ~50%
set.seed(83273)
for (i in 1:2000) {
  x$oligo[x$iteration == i] <- sim.DNAseq(18, GCfreq = 0.5)
}

# Generate all possible combinations of these motifs with primer sequences, spacers, etc.
tf.array2 <- merge(x, tf.array)

tf.array2$seq.name <- paste(tf.array2$TF, tf.array2$iteration, sep = "_")


# Assemble sequence to test: 4 rim bases of the TF motifs and the spacer sequence
tf.array2$seq.text <- paste(substrRight(tf.array2$reporter, 4), 
                       tf.array2$oligo, 
                       substr(tf.array2$reporter, 1, 4), sep = "")

# Write fasta file to run on FIMO script
tf.array.export <- tf.array2 %>% select(seq.name, seq.text)
dat2fasta(x, outfile = "/DATA/usr/m.trauernicht/projects/TF_array/data/library_design/output/x.fasta")     
```


 
## Run FIMO script 
```{bash run fimo db2_9, eval = FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
# motfn=/home/m.trauernicht/mydata/data/TFDB/natoli_update_2017/20170320_pwms_selected.meme
# odir=/DATA/usr/m.trauernicht/projects/TF_array/data/library_design/output/fimo
# query=/DATA/usr/m.trauernicht/projects/TF_array/data/library_design/output/x.fasta

# nice -n 19 fimo --no-qvalue --thresh 1e-4 --verbosity 1 --o $odir $motfn $query 
```



## load fimo results
We built a TF motif matrix using -log10 transformed FIMO scores. We used this feature encoding throughout the rest of this analysis, unless otherwise stated.

```{r build tf motif matrices db2: primer_1, out.width= "100%", fig.align= "center", warning= FALSE}
# load motif Metadata --> PWM feature matrix
tib_pwm_tf_array  <- get_pwm_feature_matrix(motif_meta_fn = '/home/m.trauernicht/mydata/data/TFDB/natoli_update_2017/fc181127_curated_metadata_no_composite_filt.csv', 
                                        fimo_fn       = '/DATA/usr/m.trauernicht/projects/TF_array/data/library_design/output/fimo/fimo.tsv',
                                        db            = 2)
```


## visualize fimo results
```{r out.width= "100%", fig.align= "center", warning= FALSE, message = FALSE}
# convert to binary
tib_pwm_tf_array_binary <- tib_pwm_tf_array %>% mutate_if(is.numeric, function(x) ifelse((x==0 | is.na(x)),0,1)) 
tib_pwm_tf_array_binary_top <- tib_pwm_tf_array_binary

# compute rowsums to get cumulative binding
tib_pwm_tf_array_binary_top$binding <- rowSums(tib_pwm_tf_array_binary_top[,2:ncol(tib_pwm_tf_array_binary_top)])

# select only cumulative binding and id
tib_pwm_tf_array_binary_top <- tib_pwm_tf_array_binary_top %>%
  select(id,binding)

# make third column wiht only the space_id - then sum up the binding scores per space_id
tib_pwm_tf_array_binary_top$space <- gsub(".*_(.*)", "\\1", tib_pwm_tf_array_binary_top$id)
tib_pwm_tf_array_binary_top <- tib_pwm_tf_array_binary_top %>%
  select(-id)
tib_pwm_tf_array_binary_top$cum_binding <- ave(tib_pwm_tf_array_binary_top$binding, tib_pwm_tf_array_binary_top$space,
                                             FUN = sum)
tib_pwm_tf_array_binary_top <- tib_pwm_tf_array_binary_top %>%
  select(-binding)
tib_pwm_tf_array_binary_top <- unique(tib_pwm_tf_array_binary_top)

# Identify spacers that were exlcuded from fimo script (due to 0 hits)
space_id <- unique(tib_pwm_tf_array_binary_top$space)
iteration <- 1:2000
space_nohit <- iteration[! iteration %in% space_id]
# 140 possible oligos

# Select oligos without EcoRI sites (as I might want to clone with EcoRI)
remove_primer_1 <- tf.array2[grep("GAATTC", tf.array2$seq.text),]
remove_primer_1 <- unique(remove_primer_1$iteration)
space_nohit <- space_nohit[!space_nohit %in% remove_primer_1]
paste (length(space_nohit), "oligos have no TF binding and could serve as adapters")
```

## Select sequences
```{r select sequences, warning= FALSE, message = FALSE}
# Select oligos
oligos_noEco <- unique(tf.array2[tf.array2$iteration %in% space_nohit,2])

# Compute GC contents & select with 9/20 - 11/20 GC content
dna_string <- DNAStringSet(oligos_noEco)
oligos_noEco <- as.data.frame(oligos_noEco)
oligos_noEco$gc <- letterFrequency(dna_string, "GC")
oligos_noEco <- oligos_noEco[oligos_noEco$gc >= 9 & oligos_noEco$gc <= 11,]

# Primer sequence should start & end with C/G
oligos_noEco <- oligos_noEco[grep(".*[C;G]$", oligos_noEco$oligos_noEco),]
oligos_noEco <- oligos_noEco[grep("^[C;G].*$", oligos_noEco$oligos_noEco),]

# Manually select top 12 selection (based on primer qualities)
oligos_noEco <- oligos_noEco[c(1,2,4,5,6,7,9,11,12,14,15,16),]

# Determine which oligos would be good as left/right anchor
oligo_left <- substr(oligos_noEco$oligos_noEco, 1, 10)
oligo_right <- substr(oligos_noEco$oligos_noEco, 9, 18)
bridge <- merge(oligo_left, oligo_right)

## Test the GC content of left/right 10bp parts
oligo_left <- substr(oligos_noEco$oligos_noEco, 1, 10)
oligo_right <- substr(oligos_noEco$oligos_noEco, 9, 18)
oligo_left <- as.data.frame(oligo_left)
oligo_right <- as.data.frame(oligo_right)
oligo_left$id_left <- 1:12
oligo_right$id_right <- 1:12

## Compute GCs
oligo_left_string <- DNAStringSet(oligo_left$oligo_left)
oligo_left$gc <- letterFrequency(oligo_left_string, "GC")
oligo_right_string <- DNAStringSet(oligo_right$oligo_right)
oligo_right$gc <- letterFrequency(oligo_right_string, "GC")
# Manual investigation gives the following separation:
right <- c("^9$","^10$","^5$","^7$","^11$","^12$")
oligo_right <- oligo_right[grep(paste(right, collapse = "|"), oligo_right$id_right),]
left <- c("^1$","^3$","^6$","^2$","^4$","^8$")
oligo_left <- oligo_left[grep(paste(left, collapse = "|"), oligo_left$id_left),]


# Fuse oligo adapters to reporters
oligo_left_sel <- oligos_noEco[c(1,2,3,4,6,8),]
oligo_right_sel <- oligos_noEco[c(5,7,9,10,11,12),]
tf.array$oligo.left <- oligo_left_sel$oligos_noEco
tf.array$oligo.right <- oligo_right_sel$oligos_noEco
tf.array$seq.text <- paste(tf.array$oligo.left, tf.array$reporter, tf.array$oligo.right, sep = "")
tf.array <- tf.array %>% select(TF, seq.text)
for (i in unique(tf.array$TF)){
p <- paste("Length oligo - ", i, ": ", nchar(tf.array$seq.text[tf.array$TF == i]), sep = "")
print(p)
}



# Check if the constructs have one of these restriction enzyme sites (needed for downstream cloning):
## AatII,  ApaI, BbsI, BmrI, BmtI, BsaI, BsgI, BsmBI, BsmI, BspQI,  BsrDI,  BtgZI,  BtgZI,  BtsI,  FseI,  KpnI,  NsiI,  PstI, SacI, SbfI, SphI
restriction.enzyme <- c("GACGTC", "GGGCCC", "GAAGAC", "ACTGGG", "GCTAGC", "GGTCTC", "GTGCAG", "CGTCTC", "GAATGC", "GCTCTTC", "GCAATG", "GCGATG", "GCAGTG", "GGCCGGCC", "GGTACC", "ATGCAT", "CTGCAG", "GAGCTC", "CCTGCAGG", "GCATGC")

# Check if the constructs have one of these nicking enzyme sites (needed for downstream cloning)
## Nb.BsmI, Nb.BsrDI, Nb.BtsI, Nb.BbvCI
nicking.enzyme <- c("GAATGC", "GCAATG", "GCAGTG", "CCTCAGC")

restriction <- tf.array[grep(paste(restriction.enzyme, collapse = "|"), tf.array$seq.text),]
nicking <- tf.array[grep(paste(nicking.enzyme, collapse = "|"), tf.array$seq.text),]
```

# Generate bridging oligos from selected oligo adapters
```{r, warning= FALSE, message = FALSE}
# 12 adapter were chosen for 6 reporters
## Now we need to choose 10 bridging oligos, overlapping with the left 10 bp of the upstream reporter and the right 10 bp of the downstream reporter
# 4 possible orders will be created

# Generate bridging oligo combinations
oligo_left <- oligo_left %>% select(-gc)
oligo_right <- oligo_right %>% select(-gc)
bridge <- merge(oligo_left, oligo_right)
bridge$oligos <- paste(bridge$oligo_right, bridge$oligo_left, sep = "")
bridge$id <- paste(bridge$id_right, bridge$id_left, sep = "_")
bridge <- bridge %>% select(oligos, id)

# Compute GC content of bridges
bridge_string <- DNAStringSet(bridge$oligos)
bridge$gc <- letterFrequency(bridge_string, "GC")

# 4 Random orders were chosen
order1 <- c("5_3","9_4","10_6","11_8","12_2")
oligos_order1 <- bridge[grep(paste(order1, collapse = "|"), bridge$id) ,]
oligos_order1$order <- 1
order2 <- c("11_3","9_1","5_8","12_2","7_4")
oligos_order2 <- bridge[grep(paste(order2, collapse = "|"), bridge$id) ,]
oligos_order2$order <- 2
order3 <- c("7_8","12_6","11_3","9_1","5_4")
oligos_order3 <- bridge[grep(paste(order3, collapse = "|"), bridge$id) ,]
oligos_order3$order <- 3
order4 <- c("10_8","12_3","9_2","7_6","11_1")
oligos_order4 <- bridge[grep(paste(order4, collapse = "|"), bridge$id) ,]
oligos_order4$order <- 4

# bridging oligos that combine complete 6-TF reporter constructs
order1_order2_order3_order4_order1 <- c("7_6", "10_2", "10_4", "5_1")
oligos_order1_order2_order3_order4_order1 <- bridge[grep(paste(order1_order2_order3_order4_order1, collapse = "|"), bridge$id) ,]
```


## Exporting data 
```{r}
# TF reporters with primer adapters
filename <- SetFileName("_TF_array_reporters", "mt")
setwd("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/oligo_design/integrated_TF_array_designT/output/")
write.csv(tf.array, file = paste(filename,".csv", sep = ""), row.names = F)

# Bridging oligos
bridging_oligos <- rbind(oligos_order1,oligos_order2,oligos_order3,oligos_order4) %>% select(oligos, order)
filename <- SetFileName("_bridging_oligos", "mt")
setwd("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/oligo_design/integrated_TF_array_designT/output/")
write.csv(bridging_oligos, file = paste(filename,".csv", sep = ""), row.names = F)
```

# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

