---
title: "Golden Gate TF reporter assembly - Sequence design"
author: "Max Trauernicht"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    theme: journal #cerulean
    highlight: monochrome
    toc: true
    toc_float: true
    code_folding: show
  editor_options:
    chunk_output_type: console
---

# knitr document van Steensel lab

# Introduction
Here, I will document how I designed the sequences for the tandem TF reporter array. The goal of the Golden Gate cloning is to assemble up to 15 different TF reporters. To do so, I will order single-stranded oligos in a pool, which can be used for PCR. This PCR adds overhangs that include typeIIs restriction enzyme sites. The primer design to add the overhangs will be done with NEBs Golde Gate Assembly Tool (https://goldengate.neb.com/). I will design 2 different methods for the assembly:
  **I. TF reporters with each the same or a unique primer annealing site**
  **II. TF reporters with BsaI annealing sites at the ends ordered as dsDNA**

**Additional design details:**
- 5' and 3' primer sequence as described above
- 4 TF motifs
- choose most optimal reporters from episomal TF reporter screen data
- Illumina primer adapter sequence after minP (same as before)
- 12bp barcode (as before)
- spacing after barcode - ~60 bp: inactive spacing (~20bp) + 2 primer annealing sites (RT & consecutive PCR)
- pA site: minimal pA site from TRIP paper
- Insulators can be added later


**Assembly controls:**
- 10-15 random shuffled reporters -> non-repetitive control

**TF reporters included:**
- Stat3
- Trp53
- Rara
- Gr
- Random (negative control)
- Nfya (positive control)
- Klf4
- Tfcp2l1
- Creb1
- Elk1
- Nfkb1
- Tcf7l2
- Nfe2l2
- Sox2/Oct4/Nanog?

**Vector backbone:**
- p101 - without GFP

## Description of Data
How to make a good rendering table: 
```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
| column1	|	column2	|	column3	|	
|----|----|----|
|1	|	2	|	3	|	
|a	|	b	|	c	|	
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

# Data processing
## Path, Libraries, Parameters and Useful Functions
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(ggplot2)
library(seqinr)
library(seqLogo)
library(universalmotif)
library(Biostrings)
library(SimRAD)
library(gtools)
library(DNABarcodes)
library(phylotools)
library(ape)
library(magrittr)
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(heatmaply)
library(pheatmap)
library(tibble)
library(ggseqlogo)
library(RColorBrewer)
library(stringi)
```

### Custom functions
Functions used thoughout this script.
```{r}
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}

# Function to substring the right part of the motif
substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

# Function to load PWM matrix
get_pwm_feature_matrix <- function(motif_meta_fn, fimo_fn, db = 2) {

  # validate args
  valid_dbs <- 1:2
  if(!db %in% valid_dbs)
    stop('Invalid db (database version). Please use db=1 (maintained for backward compatibility only) or db=2')

  # db=1 is maintained for backward compatibility only
  if(db == 1) {

    # read in motif metadata
    motif_meta    <- read.csv(motif_meta_fn)

    # check whether motif metadata contain essential annotations
    if(!all(c('PWM.ID', 'Cognate.TF') %in% colnames(motif_meta))) {
      message('The motif metadata file does not contain the essential columns PWM.ID and Cognate.TF')
    }

    motif_minimal <- motif_meta[, c('PWM.ID', 'Cognate.TF')]

    # load fimo output --> extract motif id, sequence id and p-value
    df <- read.table(fimo_fn)
    df <- df[, c(1, 2, 7)]

    colnames(df) <- c('PWM.ID', 'seqid', 'pval')

    # add TF id
    df <- merge(df, motif_minimal, by = 'PWM.ID')

    # group motif hits by sequence id
    l <- split(df, df[['seqid']])

    # multiple PWM and multiple hits possible. Reduce hits to one per TF, keeping best p-val only
    l <- lapply(l, function(x) {
      x_by_tf <- split(x, x[['Cognate.TF']], drop = TRUE)
      x_by_tf <- lapply(x_by_tf, function(y) y[which.min(y$pval), ])
      do.call('rbind', x_by_tf)
    })

    # initialize feature matrix
    n_tf          <- motif_minimal[['Cognate.TF']] %>%
      unique %>%
      length
    n_seq         <- length(l)
    pwm           <- matrix(1, nrow = n_seq, ncol = n_tf)
    colnames(pwm) <- (motif_minimal[['Cognate.TF']] %>% unique)

    # replace :: from names of composite motifs
    colnames(pwm) <- str_replace_all(colnames(pwm), '::', '_')

    # fill in feature matrix
    for(i in 1 : n_seq) {
      pwm[i, l[[i]][['Cognate.TF']]] <- l[[i]]$pval
    }

    # -log10 transform
    pwm           <- -1 * log10(pwm)

    # coerce to tib and return
    tib_fimo <- as_data_frame(pwm) %>%
      mutate(id = names(l))
      dplyr::select(id, everything())

  }

  # db = 2 (default)
  else {

    # load metadata
    tib_meta    <- read_csv(motif_meta_fn) %>%
      # extract tf symbol from motif id (Cognate_TF unsafe, it can be empty) and replace :: occurrences
      mutate(tf_symbol = str_remove(ID, '_[0-9]*'),
             tf_symbol = str_replace(tf_symbol, '::', '_')) %>%
      select(motif_id = `PWM ID`, tf_symbol)

    # load fimo results
    tib_fimo <- read_tsv(fimo_fn) %>%
      # extract motif id, sequence id and p-value
      select(motif_id, sequence_name, pval = `p-value`)

    # add tf symbol to fimo results
    tib_fimo <- tib_fimo %>%
      left_join(tib_meta, by = 'motif_id') %>%
      # remove hits with missing motif id (composite pwms)
      filter(!is.na(tf_symbol))

    # select best hit for each motif and sequence
    tib_fimo <- tib_fimo %>%
      group_by(sequence_name, tf_symbol) %>%
      dplyr::slice(which.min(pval)) %>%
      ungroup()

    # spread into feature matrix
    tib_fimo <- tib_fimo %>%
      mutate(pval = -1 * log10(pval)) %>%
      select(-motif_id) %>%
      spread(key = tf_symbol, value = pval, fill = 0, drop = TRUE) %>%
      # perform cosmetics on the id
      mutate(id = sequence_name) %>%
      select(-c(sequence_name)) %>%
      select(id, everything())

  }

  return(tib_fimo)

}
```


## Data import
```{r}
tf.array <- read.csv("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/SuRE_TF_1/TF-reporter_library_design/output/mt20191205_tf-array.csv")
```



# TF array sequence design

## Selecting TF reporters
```{r}
# Make selection df
selection_df <- data.frame(TF = c("Random1", "Sp1", "Stat3", "Klf4", "Tcfcp2l1", "Rara",
                                  "Creb1", "Elk1", "Trp53", "Nfkb1", "Tcf7l2", "Nr3c1(GR)"),
                           Spacing = c("5bp", "5bp", "10bp", "10bp", "5bp", "5bp", 
                                       "5bp", "5bp", "10bp", "5bp", "5bp", "5bp"),
                           Distance = c("10bp", "10bp", "10bp", "10bp", "10bp", 
                                        "10bp", "10bp", "10bp", "10bp", "10bp", "21bp", "10bp"),
                           Promoter = c("minP", "minP", "mCMV", "minP", "minP", "minP", 
                                        "minP", "minP", "minP", "minP", "minP", "minP"),
                           Background = c("1", "1", "3", "1", "1", "1", "1", "1", 
                                          "3", "1", "1", "1"))
selection_df <- selection_df %>%
  mutate(reporter_id = paste(TF, Spacing, Distance, Promoter, Background))

tf.array <- tf.array %>%
  mutate(reporter_id = paste(TF, Spacing, Distance, Promoter, Background)) %>%
  filter(reporter_id %in% selection_df$reporter_id,
         Barcode == 4)
tf.array$TF[tf.array$TF == "Nr3c1(GR)"]  <- "Gr"

# Remove 4th Trp53 motif to shorten the reporter
tf.array$motif4[tf.array$TF == "Trp53"] <- ""
tf.array$Space3[tf.array$TF == "Trp53"] <- ""

# Add Oct4/Sox2/Nanog?

# Add 28bp spacing after barcode - this is the maximum length which allows me to stay below 350bp total length
tf.array$bc_spacer <- "CATCGTCGCATCCAAGAGGctagctaac" #same sequence as in previous library design

# Add RT primer sequence in front of pA signal 
tf.array$RT_primer <- "gccatgaagatcgagtgccgcatcaccgtggagttcgagctggtgggcggcggagagg" #same as conventional SuRE protocol

# Attach pA site (sNRP-1)
tf.array$pA <- "cttgtgactgggaaaaccctggcgtaaataaaatacgaaatg"
```


# Generate BsaI recognition sites + overhangs  
## Get overhangs from https://goldengate.neb.com/ -> golden gate utilities
```{r}
# BsaI recognition site 5'->3'
tf.array$recognition_site <- "ggctacggtctcc"

# BsaI recognition site 3'->5'
tf.array$recognition_site_r <- "ggagaccgtagcc"

# Generated overhang from goldengate site, 5'->3'
tf.array$overhang <- c("CCTA", "ATTA", "GGGA", "ACTC", "ACAG", "ACGA", "CTAA", "GCAA", "CCCG", "CATC", "AGAT", "AGCG")

# Generated overhang from goldengate site, 5'->3'
tf.array$overhang_r <- c("ATTA", "GGGA", "ACTC", "ACAG", "ACGA", "CTAA", "GCAA", "CCCG", "CATC", "AGAT", "AGCG", "TACA")


## 5'-3' overhang for vector: "TACA"
## 3'-5' overhang for vector: "TAGG"
```






```{r}
# Concatenate intermediate sequence
tf.array$reporter <- paste(tf.array$recognition_site, tf.array$overhang, tf.array$motif1, 
                           tf.array$Space1, tf.array$motif2, tf.array$Space2, tf.array$motif3, 
                           tf.array$Space3, tf.array$motif4, tf.array$Distance_seq, tf.array$Promoter_sequence,
                           tf.array$S1_primer, tf.array$barcode, tf.array$bc_spacer,
                           tf.array$RT_primer, tf.array$pA, tf.array$Primer2_seq, tf.array$overhang_r,
                           tf.array$recognition_site_r, 
                           sep = "")
tf.array_export <- tf.array %>% select(TF, reporter)

# Check length of oligos (max of 350 is allowed by IDT)
paste("Max length:", max(nchar(tf.array_export$reporter)), ", Min length:", 
      min(nchar(tf.array_export$reporter)))

# Check for BsaI restriction sites in sequence
scan <- tf.array_export$reporter[grep("GGTCTC", tf.array_export$reporter)]
paste("I found", length(scan), "BsaI sites")
```




# Add controls
```{r}
# Scrambled TF reporters
tf.scrambled <- tf.array 
tf.scrambled$reporter <- paste(tf.array$motif1, 
                           tf.array$Space1, tf.array$motif2, tf.array$Space2, tf.array$motif3, 
                           tf.array$Space3, tf.array$motif4, tf.array$Distance_seq, tf.array$Promoter_sequence,
                           tf.array$S1_primer, tf.array$barcode, tf.array$bc_spacer,
                           tf.array$RT_primer, tf.array$pA, sep = "")
set.seed(112321)
tf.scrambled$reporter_scr <- stri_rand_shuffle(tf.scrambled$reporter) # Shuffle DNA sequence
tf.scrambled$Primer1_seq <- c("CGGAGCGAACCGAGTTAG")
tf.scrambled$Primer2_seq <- c("CATCGTCGCATCCAAGAG")
tf.scrambled$seq <- paste(tf.scrambled$Primer1_seq, tf.scrambled$reporter_scr, tf.scrambled$Primer2_seq, sep = "")
tf.scrambled_export <- tf.scrambled %>% select(TF, seq)
paste("Max length:", max(nchar(tf.scrambled_export$seq)))

# Check for BsaI restriction sites in sequence
scan <- tf.scrambled_export$seq[grep("GGTCTC", tf.scrambled_export$seq)]
paste("I found", length(scan), "BsaI sites")
```






## Reporter export
```{r}
# Export TF reporter sequences
filename <- SetFileName("_tf-reporter-seq", "mt")
setwd("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/SuRE_TF_1/TF-reporter_library_design/output/")
write.csv(tf.array_export, file = paste(filename,".csv", sep = ""), row.names = F)

filename <- SetFileName("_tf-reporter-scrambled", "mt")
setwd("/DATA/usr/m.trauernicht/projects/tf_activity_reporter/data/SuRE_TF_1/TF-reporter_library_design/output/")
write.csv(tf.scrambled_export, file = paste(filename,".csv", sep = ""), row.names = F)
```




# Conclusions
```{r}
```

## Exporting potential data. 
```{r}
```

# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

